Loop-Aware Optimizations in PyPy's Tracing JIT

### 2012-08-06 talk/dls2012/paper.tex
reduce the use of the word "simple"

\reva{
You often use the word simple. While it might make sense to use it,
it exact meaning in that context remains unclear.
}

One of the advantages that tracing JIT compilers have above traditional
method-based
JITs is that their optimizers are much easier to write. Because a tracing JIT
produces only linear pieces of code without control flow joins, many
optimization passes on traces can have a very simple structure. They often
consist of one forward pass replacing operations by simpler ones or even
discarding them as they walk along it. This makes
optimization of traces very similar to symbolic execution. Also, many
difficult problems in traditional optimizers become tractable if the optimizer
does not need to deal with control flow merges.

One disadvantage of this simplicity is that such simple forward-passing
optimizers ignore the only bit of control flow they have available, which is
the fact that most traces actually represent loops. Making use of this
information is necessary to perform optimizations that take the whole loop into
account, such as loop-invariant code
motion or optimizations that improve across several iterations of the loop.
Having to deal with this property of traces complicates the optimization passes,
as a more global view of a trace needs to be considered when optimizing.

In this paper we want to address this problem by proposing a simple scheme that
makes it possible to turn optimizations using one forward pass into
optimizations that can do loop invariant code motion and similar loop-aware
improvements. Using this scheme one does not need to change the underlying
optimization much to get these advantages.

@@

Because $i_0$ is loop-invariant, the addition could be moved out of the loop.
However, we want to get this effect using our existing optimization passes
without changing them too much. Simple optimizations with one forward pass
cannot directly get this effect: They just look at the trace without taking
into account that the trace executes many times in a row. Therefore to achieve
loop-invariant code motion, we peel one iteration off the loop before running
the optimizations. This peeling gives the following trace:

@@

This scheme is quite powerful and generalizes to other optimizations than just
common subexpression elimination. It allows simple linear optimization passes to
perform loop-aware optimizations, such as loop-invariant code motion without
changing them at all. All that is needed is to peel off one iteration, then
apply simple one-pass optimizations and make sure that the necessary extra
arguments are inserted into the label of the loop itself and the jumps
afterwards.

@@

For the purpose of this paper, we are going to use a tiny interpreter for a dynamic language with
 a very simple object
model, that just supports an integer and a float type (this example has been taken from a previous paper \cite{bolz_allocation_2011}). The objects support only
one operation, \lstinline{add}, which adds two objects (promoting ints to floats in a
mixed addition). The implementation of \lstinline{add} uses classical Smalltalk-like
double-dispatching.
%These classes could be part of the implementation of a very
%simple interpreter written in RPython.
The classes can be seen in
Figure~\ref{fig:objmodel} (written in RPython).

@@

Let us now consider a simple ``interpreter'' function \lstinline{f} that uses the
object model (see the bottom of Figure~\ref{fig:objmodel}).
Simply running this function is slow, because there are lots of virtual method
calls inside the loop, two for each
call to \lstinline{add}. These method calls need to check the type of the involved
objects every iteration. In addition, a lot of objects are created
when executing that loop, many of these objects are short-lived.
The actual computation that is performed by \lstinline{f} is simply a sequence of
float or integer additions (note that \lstinline{f} does not actually terminate,
but it is still instructive to look at the produced traces).

@@

If a pure operation appears more than once in the trace with the same input
arguments, it only needs be executed the first time and then the result
can be reused for all other appearances. PyPy's optimizers can also remove
repeated heap reads if the intermediate operations cannot have changed their
value\footnote{We perform a simple type-based alias analysis to know which
writes can affect which reads. In addition writes on newly allocated objects
can never change the value of old existing ones.}.

@@

The sobel and conv3x3 benchmarks are implemented
on top of a custom two-dimensional array class.
It is
a simple straight forward implementation providing 2 dimensionall
indexing with out of bounds checks. For the C implementations it is
implemented as a C++ class. The other benchmarks are implemented in
plain C.

>>>>

One of the advantages that tracing JIT compilers have above traditional
method-based
JITs is that their optimizers are much easier to write. Because a tracing JIT
produces only linear pieces of code without control flow joins, many
optimization passes on traces can have a very simple structure: They often
consist of one forward pass replacing operations by faster ones or even
discarding them as they walk along it. This makes
optimization of traces very similar to symbolic execution. Also, many
difficult problems in traditional optimizers become tractable if the optimizer
does not need to deal with control flow merges.

One disadvantage of this simplicity is that such forward-passing
optimizers ignore the only bit of control flow they have available, which is
the fact that most traces actually represent loops. Making use of this
information is necessary to perform optimizations that take the whole loop into
account, such as loop-invariant code
motion or optimizations that improve across several iterations of the loop.
Having to deal with this property of traces complicates the optimization passes,
as a more global view of a trace needs to be considered when optimizing.

In this paper we want to address this problem by proposing a scheme that
makes it possible to turn optimizations using one forward pass into
optimizations that can do loop invariant code motion and similar loop-aware
improvements. Using this scheme one does not need to change the underlying
optimization much to get these advantages.

@@

Because $i_0$ is loop-invariant, the addition could be moved out of the loop.
However, we want to get this effect using our existing optimization passes
without changing them too much. Optimizations with one forward pass
cannot directly get this effect: They just look at the trace without taking
into account that the trace executes many times in a row. Therefore to achieve
loop-invariant code motion, we peel one iteration off the loop before running
the optimizations. This peeling gives the following trace:

@@

This scheme is quite powerful and generalizes to other optimizations than just
common subexpression elimination. It allows linear optimization passes to
perform loop-aware optimizations, such as loop-invariant code motion without
changing them at all. All that is needed is to peel off one iteration, then
apply one-pass optimizations and make sure that the necessary extra
arguments are inserted into the label of the loop itself and the jumps
afterwards.

@@

For the purpose of this paper, we are going to use a tiny interpreter for a dynamic language with
 a very small object
model, that just supports an integer and a float type (this example has been taken from a previous paper \cite{bolz_allocation_2011}). The objects support only
one operation, \lstinline{add}, which adds two objects (promoting ints to floats in a
mixed addition). The implementation of \lstinline{add} uses classical Smalltalk-like
double-dispatching.
%These classes could be part of the implementation of a very
%simple interpreter written in RPython.
The classes can be seen in
Figure~\ref{fig:objmodel} (written in RPython).

@@

Let us now consider an ``interpreter'' function \lstinline{f} that uses the
object model (see the bottom of Figure~\ref{fig:objmodel}).
Simply running this function is slow, because there are lots of virtual method
calls inside the loop, two for each
call to \lstinline{add}. These method calls need to check the type of the involved
objects every iteration. In addition, a lot of objects are created
when executing that loop, many of these objects are short-lived.
The actual computation that is performed by \lstinline{f} is simply a sequence of
float or integer additions (note that \lstinline{f} does not actually terminate,
but it is still instructive to look at the produced traces).

@@

If a pure operation appears more than once in the trace with the same input
arguments, it only needs be executed the first time and then the result
can be reused for all other appearances. PyPy's optimizers can also remove
repeated heap reads if the intermediate operations cannot have changed their
value\footnote{We perform a type-based alias analysis to know which
writes can affect which reads \cite{XXX}. In addition writes on newly allocated objects
can never change the value of old existing ones.}.

@@

The sobel and conv3x3 benchmarks are implemented
on top of a custom two-dimensional array class.
It is
a straightforward implementation providing 2 dimensional
indexing with out of bounds checks. For the C implementations it is
implemented as a C++ class. The other benchmarks are implemented in
plain C.
